Symbolic Weighted Language Models, Quantitative Parsing and Verification\\ over Infinite Alphabets

### 2021-10-15 main.tex
submission FoSSaCS

% comments in margin with \todo command
\usepackage[colorinlistoftodos]{todonotes} % option disable to hide todos
% command todo : options
% - noline: no line connecting the note with the place in the text
%   where the note occurs in the latex code
% - fancyline : curved arrow from note to inssertion point
% - inline: place a todonote inside the text instead of in the margin
% - author=name
\newcommand{\florent}[1]{\todo[noline,size=\tiny,color=yellow!40]{#1}}
%\renewcommand{\florent}[1]{} % uncomment for final
\newcommand{\lydia}[1]{\todo[noline,size=\tiny,color=red!30]{#1}}
%\renewcommand{\lydia}[1]{} % uncomment for final
\newcommand{\philippe}[1]{\todo[noline,size=\tiny,color=green!30]{#1}}
%\renewcommand{\philippe}[1]{} % uncomment for final
\newcommand{\reviews}[1]{\todo[noline,size=\tiny,color=blue!30]{#1}}
%\renewcommand{\reviews}[1]{} % uncomment for final

@@

We follow the approach of~\cite{Mohri03ijfcs} for the computation of distances
between words and languages
and extend it to infinite alphabets.
% with models of symbolic weighted automata and transducers.
%
The models presented in this section generalize weighted automata and transducers~\cite{Droste09handbook}
%over finite alphabets, see \eg~\cite{Mohri03ijfcs},
by labeling each transition with a weight function
that takes the input and output symbols as parameters.
These functions generalize the guards 
of symbolic automata~\cite{Veanes12symbolic,dAntoniVeanes17CAV,dAntoni21CACM},
from Boolean domains to commutative semirings.
%
They are particular cases of the very general model of 
Weighted Symbolic Automata with Data Storage~\cite{Herrmann16dlt}. 
\florent{review 3: = classical transducers}

@@

\begin{definition}
\label{def:transducer} \label{def:SWT}

@@

@@

%
We recall that, by convention (Section~\ref{sec:prelim}),
an empty sum with $\bigoplus$ is equal to~$\zero$.
%
Intuitively,  a transition $\wei_{ij}(q, a, b, q')$ is interpreted as follows:
when reading $a$ and $b$ in the input and output words,
increment the current position in the input word if and only if $i = 1$,
and in the output word iff $j = 1$, %(otherwise, do not change it),
and change state from $q$ to $q'$.
When $a = \varepsilon$ (resp. $b = \varepsilon$), the current symbol
in the input (resp. output) is not read.
%
%In contrast with the models of weighted transducers over finite alphabets~\cite{Mohri03ijfcs},
%the input and output symbols at current positions are always read by transitions,
%even when they do not change the reading position the head's position.
%This is an important feature in the case of an infinite alphabet in
%order to compare input and output symbols.
%which cannot be stored in the finite memory of the transducer.
%
Since~$\zero$ is absorbing for~$\otimes$ in~$\Semiring$,
one term $\wei_{ij}(q, a, b, q'')$ equal to $\zero$ in the above expression
will be ignored in the sum, meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and $b$.
This is analogous to the case of a transition's guard not satisfied by $\<a, b>$ for
symbolic transducers~\cite{Veanes12transducers}.

@@

\begin{example}\label{ex:SWT}
We build a small \SWT over the alphabets $\Sigma$ and $\Deltai$
of Ex.~\ref{ex:running} and~\ref{distance-time},
%weighted transducer model
with two states $q_0$ and $q_1$, that
calculates the temporal distance between an input performance in~$\Sigma^*$
and the subsequence
of $\Deltai$ events in a score.
%the performance $I$ and
%the score $O$ of Example~\ref{ex:running}.
%
Given  a performed event $\mu$
and the corresponding notated event $\nu$
(\eg MIDI pitch 69 and note $\mbox{\footnotesize A4}$),
the weight  computed by the \SWT is the time distance between both,
as modeled by transitions $\wei_{11}$ below.
%
The continuation symbol $'-'$
(met for instance in \emph{ties}
$\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
or \emph{dots} \musQuarterDotted{})
%that represents continuation
is  skipped with no cost (transitions $\wei_{01}$).
%Likewise, the markups are ignored.
\[
\begin{array}{lclclcll}
\wei_{11}(q_0, \mu\tstamp{\tau}, \nu\tstamp{\tau'}, q_0) & = &
     \delta(\mu\tstamp{\tau}, \nu\tstamp{\tau'}) & \quad & %|\tau' - \tau| & \quad &
\wei_{11}(q_1, \mu\tstamp{\tau}, \nu\tstamp{\tau'}, q_0) & = &
     \delta(\mu\tstamp{\tau}, \nu\tstamp{\tau'}) &
     \mbox{if~} \nu \neq - \\ %|\tau' - \tau| & \quad &
\wei_{01}(q_0, \varepsilon, \mathsf{-}\tstamp{\tau'}, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \mathsf{-}\tstamp{\tau'}, q_0) & = & \one\\
\wei_{10}(q_0, \mu\tstamp{\tau}, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also want to take performing errors into account,
since a performer could, for example, play an unwritten extra note.
%\lydia{reformulated this sentence}
%
%Reading such a note is modelled by
The transition $\wei_{10}$,
with a fixed weight value $\alpha \in \Semiring$,
switches from state $q_0$ (normal) to $q_1$ (error)
when reading an extra note $\mu$.
%\philippe{Comprends pas cette phrase}
%\florent{$\to$ reformulation}
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.
%\florent{dernière phrase utile?}
%This \SWT is capable of evaluating the differences between a score and a performance,
%all the while ensuring that performance errors are plausible.
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

@@

%\subsection{Properties}
\noindent
The following property will be useful for
symbolic weighted parsing (Section~\ref{sec:parsing}).

@@

\section{SW Visibly Pushdown Automata and Best-Search}
\label{section:SWVPA}\label{sec:SWVPA}
We consider now a language model generalizing symbolic \VPA
(\SVPA~\cite{dAntonyAlur14SVPDA},
 themselves generalizing \VPA~\cite{AlurMadhusudan09nested} to infinite alphabets)
from Boolean semirings to arbitrary semiring domains.
It is also a particular case of 
Weighted Symbolic Automata with Data Storage~\cite{Herrmann16dlt} for a special 
type of data storage.
It associates to every nested word over an infinite alphabet a weight value
in a semiring.
Nested words can describe structures of labeled trees.
In the context of parsing, they will be useful to
represent AST % parse trees.
(see Section~\ref{sec:parsing} and Appendix~\ref{app:trees}). % parse trees. 
%\florent{see \S 5 and App.A}

@@

\begin{example}\label{ex:nested-word}
In the nested score representation $O \in \Delta^*$ in Ex.~\ref{ex:running},
$\Deltai$ corresponds to timed notes and continuations,
%the set of music events of the form $<\nu, \tau>$
%where $\nu$ is either a note (e.g. A4), the continuation symbol $-$, or a
%rest symbol R.
and $\Deltac$ and $\Deltar$ contain respectively opening and closing parentheses.
%are used to encode the score structure:
%Let $\Deltac = \{\langle, \prec\}$ and  $\Deltar = \{\rangle, \succ\}$ where $\langle$
%(resp. $\rangle$)
%and $\prec$ (resp. $\succ$) correspond to start/end of measures, and start/end of
%tuplets. Then  \includegraphics[scale=0.20]{pictures/score5.png}
%is encoded with the nested word $\langle \prec 0, \frac{3}{4}, \frac{7}{8} \succ \rangle %\langle \prec 1, \frac{4}{3}, \frac{5}{3}  \succ \rangle$
Another example is the other candidate
\includegraphics[scale=0.35,trim=0 5mm 0 0]{pictures/ex2.pdf}
of transcription of~$I$,
linearized into
$O' =$
$\ccall{\mathsf{m}}\tstamp{[0,1]}$,
$\ccall{2}\tstamp{[0,1]}$,
$\mbox{\footnotesize A4}\tstamp{0}$,
$\ccall{2}\tstamp{[\frac{1}{2},1]}$,
$-\tstamp{\frac{1}{2}}$,
$\mbox{\footnotesize B4}\tstamp{\frac{3}{4}}$,
$\creturn{2}\tstamp{1}$,
$\creturn{2}\tstamp{1}$,
$\ccall{\mathsf{m}}\tstamp{[1,2]}$,
$\ccall{3}\tstamp{[1,2]}$,
$\mbox{\footnotesize `C$\sharp$5`}\tstamp{1}$,
$\mbox{\footnotesize D5}\tstamp{1}$,
$\mbox{\footnotesize E5}\tstamp{\frac{4}{3}}$,
$\mbox{\footnotesize F5}\tstamp{\frac{5}{3}}$,
$\creturn{3}\tstamp{2}$,
$\creturn{\mathsf{m}}\tstamp{2}$,
$\creturn{\mathsf{m}}\tstamp{2}$
(see also Fig.~\ref{fig:score-tree}).
The symbol between quotes $\mbox{\footnotesize `C$\sharp$5`}$
represents an \emph{appogiatura}, %(grace-note),
\ie an ornamental note with theoretical duration~0.
\endex
\end{example}

\begin{definition}
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA)
over  $\Delta = \Deltai \uplus \Deltac \uplus \Deltar$, $\Semiring$ and $\bar\Phi$
is a tuple $A = \< Q, P, \init, \bar\wei, \final >$,
where $Q$ is a finite set of states,
$P$ is a finite set of stack symbols,
$\mathsf{in} : Q \to \Semiring$
(respectively $\mathsf{out} : Q \to \Semiring$)
are functions defining the weight for entering
(respectively leaving) a state,
and $\bar\wei$ is a sextuplet composed of the transition functions :
$\weii : Q \times P \times Q \to \Phici$,
$\weiei : Q \times Q \to \Phii$,
$\weic : Q \times P \times Q \times P \to \Phicc$,
$\weiec : Q \times P \times Q \to \Phic$,
$\weir : Q \times P \times Q \to \Phicr$,
$\weier : Q \times Q \to \Phir$.
%and
%$\weiex : Q \times Q \to \Phix$
%with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$.
\end{definition}
%
As in Section~\ref{section:transducer},
we extend the above transition functions as follows
for every $q, q' \in Q$, $p \in P$,
$a \in \Deltai$,
$\call{c} \in \Deltac$,
$\return{r} \in \Deltar$,
overloading their names: % for simplicity:
\[
\begin{array}{lll}
\weii: Q \times [\Deltac \times P] \times \Deltai \times Q \to \Semiring &
\weii(q, c, p, a, q') = \eta_\mathsf{ci}(c, a) &
\mathrm{where~} \eta_\mathsf{ci} = \weii(q, p, q'),\\
%
\weiei: Q \times \Deltai \times Q \to \Semiring &
\weiei(q, a, q') = \phi_\mathsf{i}(a) &
\mathrm{where~} \phi_\mathsf{i} = \weiei(q, q').\\
%
\weic: Q \times [\Deltac \times P] \times  [\Deltac \times P] \times Q \to \Semiring &
\weic(q, c, p, \call{c'}, p', q') = \eta_\mathsf{cc}(c, \call{c'}) &
\mathrm{where~} \eta_\mathsf{cc} = \weic(q, p, p', q'),\\
%
\weiec: Q \times [\Deltac \times P] \times Q \to \Semiring &
\weiec(q, {\call{c}}, p', q') = \phi_\mathsf{c}({\call{c}}) &
\mathrm{where~} \phi_\mathsf{c} = \weiec(q, p, q').\\
%
\weir: Q \times [\Deltac \times P] \times \Deltar \times Q \to \Semiring &
\weir(q, {\call{c}},  p, {\return{r}}, q') = \eta_\mathsf{cr}({\call{c}},  {\return{r}}) &
\mathrm{where~} \eta_\mathsf{cr} = \weir(q, p, q'),\\
%
\weier: Q \times \Deltar \times Q \to \Semiring &
\weier(q, {\return{r}}, q') = \phi_\mathsf{r}({\return{r}}) &
\mathrm{where~} \phi_\mathsf{r} = \weier(q, q').\\
\end{array}
\]
The tuples in the definition domains of the above functions 
are also called \emph{transitions} of the \SWVPA.
We denote by $\src(\tau)$ (respectively $\snd(\tau)$) the first (respectively second) 
state component of a transition~$\tau$.
%
\noindent
%The intuition is the following for the above transitions.
Intuitively, $\weiei$, $\weiec$, and $\weier$ describe the cases where the stack is empty.
The transitions of $\weii$ and $\weiei$ both read an internal symbol~$a$ 
and change state from $q$ to $q'$, without changing the stack.
Moreover, $\weii$ reads a pair made of
${\call{c}} \in \Deltac$ and $p \in P$ on the top of the stack
($c$ is compared to $a$ by the weight function $\eta_\mathsf{ci} \in \Phici$).
%
\noindent
The transitions of $\weic$ and $\weiec$ read the call symbol $\call{c}'$,
push it to the stack along with $p'$, and change state from $q$ to to $q'$.
Moreover, $\weic$ reads ${\call{c}}$ and $p$ at the top of the stack
($c$ is compared to $c'$).
%and $\weiec$ applies iff the stack is empty.
%
\noindent
Finally, $\weir$ and $\weier$ read the return symbol $\return{r}$, and change state from $q$ to to $q'$.
Moreover, $\weir$ reads and pops from stack a pair made of $\call{c}$ and $p$
($\call{c}$ is compared to $\return{r}$).
%and $\weier$ applies iff the stack is empty.
%In this case, the weight function $\phi_\mathsf{r}$
%computes a value of matching between the call and return symbols $c$ and $r$.
%This value might be set to $\zero$ in order to express that the symbols do not match.

@@

\begin{example} \label{ex:SWVPA}
\reviews{revision ex. needed. see rev. 3) and app. $\weic$, $\weir$, $\weii$ wrong.}
We consider a \SWVPA over the alphabet of Example~\ref{ex:nested-word} and with $P = Q$,
that expresses a weight related to the music notation,
or more precisely to its structural complexity.
Given a set of equivalent representations, we aim at choosing the simpler one, \ie the one with the smallest weight.

\lydia{We should find a better variable name than j... it does not render well in fractions}
Let's say the top of the stack is a tuple of the call~$\ccall{n}\tstamp{[\tau, \tau+\ell]}$ and a state $q$. Let's now consider a new call transition starting in state $q_{\frac{j}{n}}$, that is the $j$-th state of the $n$ sub-intervals of duration~$\frac{\ell}{n}$.
We thus read a new time-division symbol ~$\ccall{d}$ 
and compute the weight of the transition from $q_{\frac{j}{n}}$ to $q_{\frac{1}{d}}$, 
the first state after the new~$\ccall{d}$, reading on the top of the stack the pair
$\<\ccall{n}\tstamp{[\tau, \tau+\ell]}, q>$, 
and pushing $\<\ccall{d}\tstamp{\iota}, q_{\frac{j+1}{c}}>$ on top, or:
\(
\weic\bigl(q_{\frac{j}{n}},
           \ccall{n}\tstamp{[\tau, \tau+\ell]}, q,
           \ccall{d}\tstamp{\iota}, q_{\frac{j+1}{c}},
           q_{\frac{1}{d}} \bigr) = \alpha_d
\).
\lydia{is it ok to explain precisely this computation and not the others ? is it redundant with the explaniation before equation (5) ?}
\noindent
Reading the $k$-th musical event $\mu$ in this current sub-interval is computed with:
\(
\weii\bigl(q_{\frac{k}{d}},
           \ccall{d}\tstamp{[\tau, \tau+\ell]}, q_{\frac{j}{n}},
           \mu\tstamp{\tau + \frac{j\ell}{d}}, q_{\frac{k+1}{d}} \bigr) = \alpha_\mu
\).
Finally, this sub-interval will end after reading a return symbol $\creturn{d}$ and compute:
\(
\weir\bigl(q_{\frac{d}{d}},
           \ccall{d}\tstamp{[\tau, \tau+\ell]}, q_{\frac{i+1}{c}},
           \,\creturn{d}\tstamp{\tau + \ell}, q_{\frac{i+1}{c}} \bigr) = \one
\).
\noindent
We described earlier the special cases where the stack is empty, which applies for example when reading the first measure 
for which we only push $\<\ccall{\mathsf{m}}\tstamp{[0, 1]}, q_{\frac{1}{1}}>$ in the stack:
\(
\weiec\bigl(q_{\frac{1}{1}},
           \ccall{\mathsf{m}}\tstamp{[0, 1]}, q_{1/1},
           q_{1/1} \bigr) = \one
\).
In comparison, any other new measure would compute instead:
\(
\weic\bigl(q_{\frac{1}{1}},
          \ccall{\mathsf{m}}\tstamp{[\tau-1, \tau]}, q_{\frac{1}{1}},
          \ccall{\mathsf{m}}\tstamp{[\tau, \tau+1]}, q_{\frac{1}{1}},
          q_{\frac{1}{1}} \bigr) = \one
\).

@@

\newpage

>>>>

% comments in margin with \todo command
\usepackage[colorinlistoftodos]{todonotes} % option disable to hide todos
% command todo : options
% - noline: no line connecting the note with the place in the text
%   where the note occurs in the latex code
% - fancyline : curved arrow from note to inssertion point
% - inline: place a todonote inside the text instead of in the margin
% - author=name
\newcommand{\florent}[1]{\todo[noline,size=\tiny,color=yellow!40]{#1}}
\renewcommand{\florent}[1]{} % uncomment for final
\newcommand{\lydia}[1]{\todo[noline,size=\tiny,color=red!30]{#1}}
\renewcommand{\lydia}[1]{} % uncomment for final
\newcommand{\philippe}[1]{\todo[noline,size=\tiny,color=green!30]{#1}}
\renewcommand{\philippe}[1]{} % uncomment for final
\newcommand{\reviews}[1]{\todo[noline,size=\tiny,color=blue!30]{#1}}
\renewcommand{\reviews}[1]{} % uncomment for final

@@

We follow the approach of~\cite{Mohri03ijfcs} for the computation of distances
between words and languages
and extend it to infinite alphabets.
% with models of symbolic weighted automata and transducers.
%
The models presented in this section are weighted automata 
and transducers~\cite{Droste09handbook}
%over finite alphabets, see \eg~\cite{Mohri03ijfcs},
with transitions labeled by weight functions
that take the input and output symbols as parameters.
These functions generalize the guards 
of symbolic automata~\cite{Veanes12symbolic,dAntoniVeanes17CAV,dAntoni21CACM},
from Boolean domains to commutative semirings.
%
These models are also particular cases of the very general model of 
Weighted Symbolic Automata with Data Storage~\cite{Herrmann16dlt}. 
%\florent{review 3: = classical transducers}

@@

\begin{definition}\label{def:transducer} \label{def:SWT}

@@

Note that the SW-transducers of Definition~\ref{def:transducer} %do not form a new model, 
are actually classical weighted transducers where the alphabets are permitted to be infinite.

@@

%
We recall that, by convention (Section~\ref{sec:prelim}),
an empty sum with $\bigoplus$ is equal to~$\zero$.
%
Intuitively,  a transition $\wei_{ij}(q, a, b, q')$ is interpreted as follows:
when reading $a$ and $b$ in the input and output words,
increment the current position in the input word if and only if $i = 1$,
and in the output word iff $j = 1$, %(otherwise, do not change it),
and change state from $q$ to $q'$.
When $a = \varepsilon$ (resp. $b = \varepsilon$), the current symbol
in the input (resp. output) is not read.
%
%In contrast with the models of weighted transducers over finite alphabets~\cite{Mohri03ijfcs},
%the input and output symbols at current positions are always read by transitions,
%even when they do not change the reading position the head's position.
%This is an important feature in the case of an infinite alphabet in
%order to compare input and output symbols.
%which cannot be stored in the finite memory of the transducer.
%
Since~$\zero$ is absorbing for~$\otimes$, 
and neutral for $\oplus$ in~$\Semiring$,
if $\wei_{ij}(q, a, b, q'')$ is equal to $\zero$
(meaning that there is no possible transition
from state $q$ into state $q'$ while reading $a$ and $b$), 
then the entire term containing this expression can be ignored in the sum.
This is analogous to the case of a transition's guard not satisfied by $\<a, b>$ for
symbolic transducers~\cite{Veanes12transducers}.

@@

\begin{example}\label{ex:SWT}
We build a small \SWT over the alphabets $\Sigma$ and $\Deltai$
of Ex.~\ref{ex:running} and~\ref{distance-time},
%weighted transducer model
with two states $q_0$ and $q_1$, that
calculates the temporal distance between an input performance in~$\Sigma^*$
and the subsequence
of $\Deltai$ events in a score.
%the performance $I$ and
%the score $O$ of Example~\ref{ex:running}.
%
Given  a performed event $\mu$
and the corresponding notated event $\nu$
(\eg MIDI pitch 69 and note $\mbox{\footnotesize A4}$),
the weight  computed by the \SWT is the time distance between both,
as modeled by transitions $\wei_{11}$ below.
%
The continuation symbol $'-'$
(met for instance in \emph{ties}
$\musQuarter\!\!\!\mathrel{\raisebox{-1.5mm}{$\smile$}}\!\!\!\musEighth$,
or \emph{dots} \musQuarterDotted{})
%that represents continuation
is  skipped with no cost (transitions $\wei_{01}$).
%Likewise, the markups are ignored.
\[
\begin{array}{lclclcll}
\wei_{11}(q_0, \mu\tstamp{\tau}, \nu\tstamp{\tau'}, q_0) & = &
     \delta(\mu\tstamp{\tau}, \nu\tstamp{\tau'}) & \quad & %|\tau' - \tau| & \quad &
     \mbox{if~} \nu \neq - \\ %|\tau' - \tau| & \quad &
\wei_{11}(q_1, \mu\tstamp{\tau}, \nu\tstamp{\tau'}, q_0) & = &
     \delta(\mu\tstamp{\tau}, \nu\tstamp{\tau'}) & &
     \mbox{if~} \nu \neq - \\ %|\tau' - \tau| & \quad &
\wei_{01}(q_0, \varepsilon, \mathsf{-}\tstamp{\tau'}, q_0) & = & \one & &
\wei_{01}(q_1, \varepsilon, \mathsf{-}\tstamp{\tau'}, q_0) & = & \one\\
\wei_{10}(q_0, \mu\tstamp{\tau}, \varepsilon, q_1) & = & \alpha & & %\multicolumn{3}{l}{\mathrm{for~all~} b \in \Delta}
\end{array}
\]
%
We also want to take performing errors into account,
since a performer could, for example, play an unwritten extra note.
%\lydia{reformulated this sentence}
%
%Reading such a note is modelled by
The transition $\wei_{10}$,
with a fixed weight value $\alpha \in \Semiring$,
switches from state $q_0$ (normal) to $q_1$ (error)
when reading an extra note $\mu$.
%\philippe{Comprends pas cette phrase}
%\florent{$\to$ reformulation}
The transitions in the second column below switch back to the normal state $q_0$.
% the metric computed by the \SWT is the smallest sum of point wise distances
% between dates of input and output events.
At last, we let $q_0$ be the only initial and final state, with
$\init(q_0) = \final(q_0) = \one$, and
$\init(q_1) = \final(q_1) = \zero$.
%\florent{dernière phrase utile?}
%This \SWT is capable of evaluating the differences between a score and a performance,
%all the while ensuring that performance errors are plausible.
%$\init(q_0, d, b) = \final(q_0, d, b) = \one$, and
%$\init(q_1, d, b) = \final(q_1, d, b) = \zero$,
%for all $d \in \Sigma$ and $b \in \Delta$.
\endex
\end{example}

@@

%\subsection{Properties}
%\noindent
%The following property will be useful for
%symbolic weighted parsing (Section~\ref{sec:parsing}).

@@

\section{SW Visibly Pushdown Automata and Best-Search}
\label{section:SWVPA}\label{sec:SWVPA}
We consider now a language model generalizing symbolic \VPA
(\SVPA~\cite{dAntonyAlur14SVPDA},
 themselves generalizing \VPA~\cite{AlurMadhusudan09nested} to infinite alphabets)
from Boolean semirings to arbitrary semiring domains.
It is also a particular case of 
Weighted Symbolic Automata with Data Storage~\cite{Herrmann16dlt} for a special 
type of data storage, and 
associates to every nested word over an infinite alphabet a weight value
in a semiring.
Nested words can describe structures of labeled trees.
In the context of parsing, they will be useful to
represent AST % parse trees.
(see Section~\ref{sec:parsing} and Appendix~\ref{app:trees}). % parse trees. 
%\florent{see \S 5 and App.A}

@@

\begin{example}\label{ex:nested-word}
In the nested score representation $O \in \Delta^*$ in Ex.~\ref{ex:running},
$\Deltai$ corresponds to timed notes and continuations,
%the set of music events of the form $<\nu, \tau>$
%where $\nu$ is either a note (e.g. A4), the continuation symbol $-$, or a
%rest symbol R.
and $\Deltac$ and $\Deltar$ contain respectively opening and closing parentheses.
%are used to encode the score structure:
%Let $\Deltac = \{\langle, \prec\}$ and  $\Deltar = \{\rangle, \succ\}$ where $\langle$
%(resp. $\rangle$)
%and $\prec$ (resp. $\succ$) correspond to start/end of measures, and start/end of
%tuplets. Then  \includegraphics[scale=0.20]{pictures/score5.png}
%is encoded with the nested word $\langle \prec 0, \frac{3}{4}, \frac{7}{8} \succ \rangle %\langle \prec 1, \frac{4}{3}, \frac{5}{3}  \succ \rangle$
Another example is the other candidate
\includegraphics[scale=0.35,trim=0 5mm 0 0]{pictures/ex2.pdf}
of transcription of~$I$,
linearized into
$O' =$
$\ccall{\mathsf{m}}\tstamp{[0,1]}$,
$\ccall{2}\tstamp{[0,1]}$,
$\mbox{\footnotesize A4}\tstamp{0}$,
$\ccall{2}\tstamp{[\frac{1}{2},1]}$,
$-\tstamp{\frac{1}{2}}$,
$\mbox{\footnotesize B4}\tstamp{\frac{3}{4}}$,
$\creturn{2}\tstamp{1}$,
$\creturn{2}\tstamp{1}$,
$\creturn{\mathsf{m}}\tstamp{1}$,
$\ccall{\mathsf{m}}\tstamp{[1,2]}$,
$\ccall{3}\tstamp{[1,2]}$,
$\mbox{\footnotesize `C$\sharp$5`}\tstamp{1}$,
$\mbox{\footnotesize D5}\tstamp{1}$,
$\mbox{\footnotesize E5}\tstamp{\frac{4}{3}}$,
$\mbox{\footnotesize F5}\tstamp{\frac{5}{3}}$,
$\creturn{3}\tstamp{2}$,
$\creturn{\mathsf{m}}\tstamp{2}$
(see also Fig.~\ref{fig:score-tree}).
The symbol between quotes $\mbox{\footnotesize `C$\sharp$5`}$
represents an \emph{appogiatura}, %(grace-note),
\ie an ornamental note with theoretical duration~0.
\endex
\end{example}

\begin{definition} 
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA)
over  $\Delta = \Deltai \uplus \Deltac \uplus \Deltar$, $\Semiring$ and $\bar\Phi$
is a tuple $A = \< Q, P, \init, \bar\wei, \final >$,
where $Q$ is a finite set of states,
$P$ is a finite set of stack symbols,
$\mathsf{in} : Q \to \Semiring$
(respectively $\mathsf{out} : Q \to \Semiring$)
are functions defining the weight for entering
(respectively leaving) a state,
and $\bar\wei$ is a sextuplet composed of the transition functions :
$\weii : Q \times P \times Q \to \Phici$,
$\weiei : Q \times Q \to \Phii$,
$\weic : Q \times P \times Q \times P \to \Phicc$,
$\weiec : Q \times P \times Q \to \Phic$,
$\weir : Q \times P \times Q \to \Phicr$,
$\weier : Q \times Q \to \Phir$.
%and
%$\weiex : Q \times Q \to \Phix$
%with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$.
\end{definition}
%
%As in Section~\ref{section:transducer},
We extend the above transition functions as follows
for every $q, q' \in Q$, $p \in P$,
$a \in \Deltai$,
$\call{c} \in \Deltac$,
$\return{r} \in \Deltar$,
overloading their names: % for simplicity:
\[
\begin{array}{lcl}
\weii: Q \times [\Deltac \times P] \times \Deltai \times Q \to \Semiring & &
\weii(q, c, p, a, q') = \eta_\mathsf{ci}(c, a)\\ 
\multicolumn{3}{r}{\mathrm{where~} \eta_\mathsf{ci} = \weii(q, p, q'),}\\
%
\weiei: Q \times \Deltai \times Q \to \Semiring & &
\weiei(q, a, q') = \phi_\mathsf{i}(a)\\
\multicolumn{3}{r}{\mathrm{where~} \phi_\mathsf{i} = \weiei(q, q'),}\\
%
\weic: Q \times [\Deltac \times P] \times  [\Deltac \times P] \times Q \to \Semiring \  & &
\weic(q, c, p, \call{c'}, p', q') = \eta_\mathsf{cc}(c, \call{c'})\qquad\\
\multicolumn{3}{r}{\mathrm{where~} \eta_\mathsf{cc} = \weic(q, p, p', q'),}\\
%
\weiec: Q \times [\Deltac \times P] \times Q \to \Semiring & &
\weiec(q, {\call{c}}, p', q') = \phi_\mathsf{c}({\call{c}})\\
\multicolumn{3}{r}{\mathrm{where~} \phi_\mathsf{c} = \weiec(q, p, q'),}\\
%
\weir: Q \times [\Deltac \times P] \times \Deltar \times Q \to \Semiring & &
\weir(q, {\call{c}},  p, {\return{r}}, q') = \eta_\mathsf{cr}({\call{c}},  {\return{r}})\\
\multicolumn{3}{r}{\mathrm{where~} \eta_\mathsf{cr} = \weir(q, p, q'),}\\
%
\weier: Q \times \Deltar \times Q \to \Semiring & &
\weier(q, {\return{r}}, q') = \phi_\mathsf{r}({\return{r}})\\
\multicolumn{3}{r}{\mathrm{where~} \phi_\mathsf{r} = \weier(q, q')}.\\
\end{array}
\]
The tuples in the definition domains of the above functions 
are also called \emph{transitions} of the \SWVPA.
We denote by $\src(\tau)$ (respectively $\snd(\tau)$) the first (respectively second) 
state component of a transition~$\tau$.
%
\noindent
%The intuition is the following for the above transitions.
Intuitively, $\weiei$, $\weiec$, and $\weier$ describe the cases where the stack is empty.
The transitions of $\weii$ and $\weiei$ both read an internal symbol~$a$ 
and change state from $q$ to $q'$, without changing the stack.
Moreover, $\weii$ reads a pair made of
${\call{c}} \in \Deltac$ and $p \in P$ on the top of the stack
($c$ is compared to $a$ by the weight function $\eta_\mathsf{ci} \in \Phici$).
%
\noindent
The transitions of $\weic$ and $\weiec$ read the call symbol $\call{c}'$,
push it to the stack along with $p'$, and change state from $q$ to to $q'$.
Moreover, $\weic$ reads ${\call{c}}$ and $p$ at the top of the stack
($c$ is compared to $c'$).
%and $\weiec$ applies iff the stack is empty.
%
\noindent
Finally, $\weir$ and $\weier$ read the return symbol $\return{r}$, and change state from $q$ to to $q'$.
Moreover, $\weir$ reads and pops from stack a pair made of $\call{c}$ and $p$
($\call{c}$ is compared to $\return{r}$).
%and $\weier$ applies iff the stack is empty.
%In this case, the weight function $\phi_\mathsf{r}$
%computes a value of matching between the call and return symbols $c$ and $r$.
%This value might be set to $\zero$ in order to express that the symbols do not match.

@@

\begin{example} \label{ex:SWVPA}
\reviews{revision ex. needed. see rev. 3) and app. $\weic$, $\weir$, $\weii$ wrong.}
We consider a \SWVPA over the alphabet of Example~\ref{ex:nested-word} and with $P = Q$,
that expresses a weight related to the music notation,
or more precisely to its structural complexity.
Given a set of equivalent representations, 
we aim at choosing the simplest one, \ie the one with the smallest weight.

\lydia{We should find a better variable name than j... it does not render well in fractions}
Let us assume that the top of the stack is a tuple 
made of the call symbol~$\ccall{n}\tstamp{[\tau, \tau+\ell]}$ 
and the state~$q$. 
Let us now consider a new call transition starting in state $q_{\frac{j}{n}}$, that is the $j$-th state of the $n$ sub-intervals of duration~$\frac{\ell}{n}$.
We thus read a new time-division symbol ~$\ccall{d}$ 
and compute the weight of the transition from $q_{\frac{j}{n}}$ to $q_{\frac{1}{d}}$, 
the first state after the new~$\ccall{d}$, reading on the top of the stack the pair
$\<\ccall{n}\tstamp{[\tau, \tau+\ell]}, q>$, 
and pushing $\<\ccall{d}\tstamp{\iota}, q_{\frac{j+1}{c}}>$ on top, or:
\(
{\weic\bigl(q_{\frac{j}{n}},
           \ccall{n}\tstamp{[\tau, \tau+\ell]}, q,
           \ccall{d}\tstamp{\iota}, q_{\frac{j+1}{c}},
           q_{\frac{1}{d}} \bigr) = \alpha_d}
\).
\lydia{is it ok to explain precisely this computation and not the others ? is it redundant with the explaniation before equation (5) ?}
\noindent
Reading the $k$-th musical event $\mu$ in this current sub-interval is computed with:
\(
\weii\bigl(q_{\frac{k}{d}},
           \ccall{d}\tstamp{[\tau, \tau+\ell]}, q_{\frac{j}{n}},
           \mu\tstamp{\tau + \frac{j\ell}{d}}, q_{\frac{k+1}{d}} \bigr) = \alpha_\mu
\).
Finally, this sub-interval will end after reading a return symbol $\creturn{d}$ and compute:
\(
{\weir\bigl(q_{\frac{d}{d}},
           \ccall{d}\tstamp{[\tau, \tau+\ell]}, q_{\frac{i+1}{c}},
           \,\creturn{d}\tstamp{\tau + \ell}, q_{\frac{i+1}{c}} \bigr) = \one}
\).
\noindent
We described earlier the special cases where the stack is empty, which applies for example when reading the first measure 
for which we only push $\<\ccall{\mathsf{m}}\tstamp{[0, 1]}, q_{\frac{1}{1}}>$ in the stack:
\(
\weiec\bigl(q_{\frac{1}{1}},
           \ccall{\mathsf{m}}\tstamp{[0, 1]}, q_{1/1},
           q_{1/1} \bigr) = \one
\).
In comparison, any other new measure would compute instead:
\(
\weic\bigl(q_{\frac{1}{1}},
          \ccall{\mathsf{m}}\tstamp{[\tau-1, \tau]}, q_{\frac{1}{1}},
          \ccall{\mathsf{m}}\tstamp{[\tau, \tau+1]}, q_{\frac{1}{1}},
          q_{\frac{1}{1}} \bigr) = \one
\).

@@

%\newpage
